Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-17 BN_discretizer_paper.tex
add wine exp plots
Modify exp part

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

@@

\subsubsection{Structure is Given}

The proposed discretization method and the MDL discretization method are tested on Auto MPG data given the network network in Fig. \ref{Auto_graph_1}.
The structure was obtained by prediscretizing each continuous variable into five uniform-width intervals, where five is the median cardinality of the discrete variables, and then taking the most frequent structure from a thousand runs of K2.

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph_autocar_1}
  \end{tabular}
  \caption{The Bayesian network obtained with K2 on the prediscretized Auto MPG dataset}
  \label{Auto_graph_1}
\end{figure}

Table \ref{Auto_MPG_disc_table_1} shows the numbers of intervals after discretization, the positions of discretization edges, and log-likelihood of 10-fold cross validation. For the Bayesian network in Fig. \ref{Auto_graph_1}, the approximation only reduces runtime of proposed method and has no influence on discretization result. MDL principle discretization works not well, since it discretizes all variables with only one interval. For a Bayesian network on thie Auto MPG data set that MDL principle discretization has non-trivial result, please reference to Appendix.

\begin{table}[h]
  \centering
  \input{table_autoMPG_1.tex}
  \caption{Discretization result of Auto MPG dataset based on the graph in Fig. \ref{Auto_graph_1}. The discretization policy for the heuristic matches that for the optimal Bayesian approach. \todo{Need to explain why MDL fails}}
  \label{Auto_MPG_disc_table_1}
\end{table}

Figure \ref{auto_exp1_distr_1_3} compares the discretization result on continuous variables $A$ and $C$ with the raw data of Auto MPG data set. The color of discretized area shows the probability of a sample being in the discretized area. The black dots in Figure \ref{auto_exp1_distr_1_3} demonstrate the distribution of the raw data. As shown in the figure, the discretization edges and the resulting marginalized probability match the raw data very well.

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_auto_exp1_distr_1_3}
    \end{tabular}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$}
  \label{auto_exp1_distr_1_3}
\end{figure}

\subsubsection{Structure is Learned}

If the network structure is not known in advance, it should be learned along with the discretization process. Figure \ref{Auto_graph_2} and Table \ref{Auto_MPG_disc_table_2} show a learned discrete-valued Bayesian network and the corresponding discretization edges of each continuous variable. This result is obtained by running Algorithm \ref{Learn_DBN} fifty times and choosing the one with the highest K2 score.
\newline

 {\centering
\begin{minipage}{0.50\textwidth}
  \centering
  \scalebox{0.7}{
  \input{graph_autocar_2}}
  \captionof{figure}{The learned discrete-\newline valued Bayesian network}
  \label{Auto_graph_2}
\end{minipage}
\begin{minipage}{0.50\textwidth}
  \raggedleft
  \input{table_autoMPG_2.tex}
  \captionof{table}{The discretization edges of variables in Fig \ref{Auto_graph_2}}
  \label{Auto_MPG_disc_table_2}
\end{minipage}
}

Figure \ref{auto_exp2_distr_1_3} compares the result of Figure \ref{Auto_graph_2} and Table \ref{Auto_MPG_disc_table_2} with the raw data on variables $A$ and $C$. Again, the black dots are raw data points. Although the numbers of discretization edges of $A$ and $C$ shown in Table \ref{Auto_MPG_disc_table_2} are less than Table \ref{Auto_MPG_disc_table_1}, they still have a good capture of raw data distribution.

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_auto_exp2_distr_1_3}
    \end{tabular}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ according to Figure \ref{Auto_graph_2} and Table \ref{Auto_MPG_disc_table_2}}
  \label{auto_exp2_distr_1_3}
\end{figure}

@@

\subsubsection{Experiment 1 on Wine dataset}
\label{subsubsec:wine_exp2}

Figure \ref{Wine_graph_1} is obtained by equal-width discretizing each continuous variable into 3 intervals and running K2 1000 times. Table \ref{Wine_disc_table_1} shows a comparison between the proposed method and the MDL principle discretization method. The proposed method is still better than MDL principle discretization in terms of likelihood, although the latter returns non-trivial result in this experiment. Furthermore, the proposed method with the approximation has more intervals after discretization, thus it has a slightly better result. However, as shown in next dataset, the approximation might cut continuous variables into too many intervals. Therefore, the proposed method without the approximation should be still considered a better approach.

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_wine_1}
    \end{tabular}
  \caption{K2 structure learning with prediscretization on Wine dataset}
  \label{Wine_graph_1}
\end{figure}

\begin{table}
\centering
\scalebox{1.0}{
\input{table_wine_1.tex}
}
\caption{Discretization result of Wine dataset based on the graph Fig. \ref{Wine_graph_1}}
\label{Wine_disc_table_1}
\end{table}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_wine_2}
    \end{tabular}
  \caption{K2 structure learning with the proposed discretization method on Wine dataset}
  \label{Wine_graph_2}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepgfplotslibrary{groupplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

@@

\subsubsection{Experiment 1 on the Auto MPG dataset}

The proposed discretization method and the MDL discretization method are tested on Auto MPG data given the network network in Fig. \ref{fig:auto_graph_1}. This structure was obtained by prediscretizing each continuous variable into five uniform-width intervals, where five is the median cardinality of the discrete variables, and then taking the most frequent structure from a thousand runs of K2.

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph_autocar_1}
  \end{tabular}
  \caption{The Bayesian network obtained with K2 on the prediscretized Auto MPG dataset}
  \label{fig:auto_graph_1}
\end{figure}

Table \ref{table:auto_disc_table_1} shows the numbers of intervals after discretization, the positions of discretization edges, and log-likelihood of 10-fold cross validation. For the Bayesian network in Fig. \ref{fig:auto_graph_1}, the approximation only reduces runtime of proposed method and has no influence on discretization result. MDL principle discretization works not well, since it discretizes all variables with only one interval. For a Bayesian network on the Auto MPG data set that MDL principle discretization has non-trivial result, please reference to Appendix.

\begin{table}[h]
  \centering
  \input{table_autoMPG_1.tex}
  \caption{Discretization result of Auto MPG dataset based on the graph in Fig. \ref{fig:auto_graph_1}. The discretization policy for the heuristic matches that for the optimal Bayesian approach. \todo{Need to explain why MDL fails}}
  \label{table:auto_disc_table_1}
\end{table}

Figure \ref{fig:auto_exp1_distr_1_3} compares the discretization result on continuous variables $A$ and $C$ with the raw data of Auto MPG data set. The color of discretized area shows the probability of a sample being in the discretized area. The black dots in Figure \ref{auto_exp1_distr_1_3} demonstrate the distribution of the raw data. As shown in the figure, the discretization edges and the resulting marginalized probability match the raw data well.

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_auto_exp1_distr_1_3}
    \end{tabular}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$}
  \label{fig:auto_exp1_distr_1_3}
\end{figure}

\subsubsection{Experiment 2 on the Auto MPG dataset}

If the network structure is not known in advance, it should be learned along with the discretization process. Figure \ref{fig:auto_graph_2} show a learned discrete-valued Bayesian network and the corresponding discretization edges of each continuous variable. This result is obtained by running Algorithm \ref{alg:structure_learn} fifty times and choosing the one with the highest K2 score. The numbers next to the variables are cardinality numbers.

\begin{figure}[ht]
\centering
    \begin{tabular}{cc}
      \input{graph_autocar_2}
    \end{tabular}
  \caption{The learned discrete-valued Bayesian network from the Auto MPG dataset}
  \label{fig:auto_graph_2}
\end{figure}

 %{\centering
%\begin{minipage}{0.50\textwidth}
%  \centering
%  \scalebox{0.7}{
%  \input{graph_autocar_2}}
%  \captionof{figure}{The learned discrete-\newline valued Bayesian network}
%  \label{fig:auto_graph_2}
%\end{minipage}
%\begin{minipage}{0.50\textwidth}
%  \raggedleft
%  \input{table_autoMPG_2.tex}
%  \captionof{table}{The discretization edges of variables in Fig \ref{Auto_graph_2}}
%  \label{Auto_MPG_disc_table_2}
%\end{minipage}
%}

Figure \ref{auto_exp2_distr_1_3} compares the edges of the learned network in Figure \ref{fig:auto_graph_2} with the raw data on variables $A$ and $C$. Again, the black dots are raw data points. Although the numbers of discretization edges of $A$ and $C$ shown in Figure \ref{fig:auto_graph_2} are less than Table \ref{table:auto_disc_table_1}, they still have a good capture of raw data distribution.

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_auto_exp2_distr_1_3}
    \end{tabular}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ of the learned network in Figure \ref{fig:auto_graph_2}}.
  \label{fig:auto_exp2_distr_1_3}
\end{figure}

@@

\subsubsection{Experiment 1 on the Wine dataset}
\label{subsubsec:wine_exp1}

The Bayesian network in Figure \ref{fig:wine_graph_1} is obtained by equal-width discretizing each continuous variable into 3 intervals and running K2 1000 times. Table \ref{table:wine_disc_table_1} shows a comparison between the proposed method and the MDL principle discretization method. The proposed method is still better than MDL principle discretization in terms of likelihood, although the latter returns non-trivial result in this experiment. Note that the proposed method with the approximation has more intervals after discretization, thus it has a slightly better result. However, as shown in next dataset, the approximation might cut continuous variables into too many intervals. Therefore, the proposed method without the approximation should be still considered a better approach.

\begin{figure}[ht]
\centering
    \begin{tabular}{cc}
      \input{graph_wine_1}
    \end{tabular}
  \caption{K2 structure learning with prediscretization on Wine dataset}
  \label{fig:wine_graph_1}
\end{figure}

\begin{table}
\centering
\scalebox{1.0}{
\input{table_wine_1.tex}
}
\caption{Discretization result of Wine dataset based on the graph Fig. \ref{Wine_graph_1}}
\label{table:wine_disc_table_1}
\end{table}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_wine_exp1_distr}
    \end{tabular}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ according to Figure \ref{Auto_graph_2} and Table \ref{Auto_MPG_disc_table_2}}
  \label{wine_exp1_distr}
\end{figure}

\subsubsection{Experiment 2 on the Wine dataset}
\label{subsubsec:wine_exp2}

Figure \ref{Wine_graph_2} is the learned discrete-valued Bayesian network, which is obtained by running Algorithm \ref{alg:structure_learn} 50 times. The integer numbers next to variables are cardinality numbers of variables. By comparing Figure \ref{Wine_graph_2} with \ref{Wine_graph_1}, it shows that the the Bayesian network obtained by \ref{alg:structure_learn} has more edges than the network obtained by prediscretization and the K2 algorithm. Furthermore, the cardinalitis of discretized variables in Figure \ref{Wine_graph_2} are smaller than the result in \ref{Wine_disc_table_1}.

Figure \ref{wine_exp2_distr_5_11} shows the discretization result on variable $E$ and $K$. The discretization edge on variable $E$ is same as the edge found by the MDL principle and the proposed method in Figure \ref{wine_exp1_distr}. This indicates that some discretization edges are robust against the network structure.

\begin{figure}[ht]
    \begin{tabular}{cc}
    \centering
    \scalebox{0.7}{
      \input{graph_wine_2}}
    \end{tabular}
  \caption{The learned discrete-valued Bayesian network from the Wine dataset}
  \label{Wine_graph_2}
\end{figure}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_wine_exp2_distr_5_11}
    \end{tabular}
  \caption{Comparison of the raw data and discretization result on variable $E$ and $K$ according to Figure \ref{Auto_graph_2}}
  \label{Wine_graph_2}
\end{figure}
